A Computational Account of Default Mode Function\\ by Reinforcement Learning Theory

### 2017-03-26 paper/paper.tex
more

\subsection{Value functions and policies}
The behavior of the agent is governed by some kind of \textit{policy}, which maps states to
sampling probabilities over candidate actions to perform from this state. Starting a time $t=0$,
a policy $\pi$ generates a trajectory as follows
\begin{eqnarray*}
  \begin{split}
    S_1 &\sim p(.|S_0,A_0), \text{ collect rewards }R_0 = r(S_0, A_0)\\
    A_1 &\sim \pi(.|S_1)\\
    S_2 &\sim p(.|S_1,A_1), \text{ collect rewards }R_1 = r(S_1, A_1)\\
    \ldots\\
    A_{t} &\sim \pi(.|S_{t})\\
    S_{t+1} &\sim p(.|S_{t},A_{t}), \text{ collect rewards }R_{t} = r(S_{t}, A_{t})\\
    \ldots
  \end{split}
\end{eqnarray*}
Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative reward estimate which takes into account the future. A standard for
modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t.
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following a
policy $\pi$.
\begin{mdframed}
  \textbf{Where is ``value'' buffered in the brain ?}
  The amygdala is known to be involved in significance evaluation) and
  the nucleus accumbens is involved in reward evaluation.
  \end{mdframed}
The  goal of RL is then adapt this policy so as to  maximize the cumulative rewards.
In the definition of $G^\pi$ above, the constant $\gamma \in [0, 1]$ is the \textit{discount factor}.
Taking $\gamma = 0$   corresponds to a perfectly myopic agent who is solely concerned about their
immediate rewards and has no horizon for the future, which is not compatible with long-term goal planning
as implemented in the DMN. For there to be any learning at all, it is thus necessary that $\gamma > 0$.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. The reward importance hence decays exponentially
with distance in the future,
given that there is more uncertainty in the farsighted future.
The reward discount factor $\gamma$ can thus be seen as calibrating
the delay discounting behavior of the intelligent agent.
It is also important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
For instance, a deterministic strategy in playing rock-paper-scissors can
be easily exploited and a uniformly random choice is optimal.

The agent's goal can then be formulated as being that of finding policy $\pi$ which maximizes the
expected cumulative value of state-stat-pairs $(A_0,S_0)$, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|S_0=s,A_0=a,A_1 \sim \pi,A_2 \sim \pi,\ldots].
\end{equation}
That is, the expectated future reward over all possible trajectories in which we start at $s$ take
action $a$ and then follow the policy $\pi$ thereafter to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(S_t,A_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.

By construction, in general, maximizing the expectation in  \eqref{eq:qval} is intractable due to
the enormous number of possible trajectories the agent might need to
consider.
(If the Bellman-based computational ideas have a corresponding
neurobiological implementation, this provides a framework of how
optimal behavior could be computable in the human brain.)
One popular simplification that removes this issue is to only consider the sub-family of
\textit{deterministic policies}, which map each state to the best single best action which
should be taken at thin state, and therefore define functions from states to actions.
\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} [Sutton and Barto 1998] then takes the simple form
\begin{equation}
  Q^\pi(s, a) = \mathbb E_{s' \sim p(.|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q^\pi(s', a')]
  = \mathbb E_{s' \sim p(.|s,a)} [r(s,a) + \gamma Q^\pi(s', \pi(s'))]
  \label{eq:bellman}
\end{equation}
which provides a recursive decomposition of optimal behavior.
A complicated dynamic programming optimization can be
broken into simpler subproblems at different time points,
which has often been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value,
to guide action (i.e., improve the policy) towards a better state.
It is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

\subsection{Efficient RL with value approximation and replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
We need ways to approximate the state-action value function
which scale up to large state spaces and / or action spaces.
In state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new state $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$D \leftarrow D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ and trying to
approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q_k)$, with the output of a parametrized regression model $(s,a,r,s')
\mapsto {Q}(s, a|\theta^Q_{k})$, where $Q(.,.|\theta^Q_{k})$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q_{k+1}$.
Computing an optimal policy then corresponds to finding the parameters $\theta^Q_{k+1}$ which minimize the following mean-squared loss function
\begin{equation}
  \mathcal L(\theta^Q_{k+1})
  % = \mathbb E_{(s, a, r, s') \sim \mathcal U(D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q_{k+1}))^2]
  = \mathbb E_{(s, a, r, s') \sim \mathcal U(D)}\left[\frac{1}{2}(Q(s, a|\theta^Q_{k+1}) - y)^2\right],
  \label{eq:oracle}
\end{equation}
where
$ y = r + \gamma \max_{a' \in \mathcal A} Q(s', a'|\theta^Q_k) = r + \gamma Q(s', \pi(s')|\theta^Q_k)$ is the prediction target.
% While $y_t$ is also dependent on $\theta^Q$, this is typically ignored.
For example, a general linear model with a kernel $\phi$ would be of the
form
$${Q}(s, a|\theta^Q) = \phi(s,a)^T\theta^Q.$$
$\phi(s,a)$ would represent a high-level hand-crafted representation of the state-action $(s,a)$ into an appropriate
feature space. Such are the techniques that have been proposed by [sutton et al. xyzw, Song et al. 2016 (NIPS),
  and Parr et al. 2003 (JMLR)].
A breakthrough alternative proposed by Google DeepMind [Minh et al. 2014, Silver et al 2015] is to learn this
representation using a deep neural net, leading to the so-called \textit{Deep Q-learning} family of methods which
are by far the current state-of-the-art in RL. In such a case, $\theta^Q$
would correspond to ..

The entire model can then be effectively trained via \textit{backprop}
wherein online stochastic gradient descent is used to minimize the
variational objective \eqref{eq:oracle}.

>>>>

\subsection{Value functions and policies}
The behavior of the agent is governed by some kind of \textit{policy}, which maps states to
sampling probabilities over candidate actions to perform from this state. Starting a time $t=0$,
a policy $\pi$ generates a trajectory as follows
\begin{eqnarray*}
  \begin{split}
    S_1 &\sim p(.|S_0,A_0), \text{ collect rewards }R_0 = r(S_0, A_0)\\
    A_1 &\sim \pi(.|S_1)\\
    S_2 &\sim p(.|S_1,A_1), \text{ collect rewards }R_1 = r(S_1, A_1)\\
    \ldots\\
    A_{t} &\sim \pi(.|S_{t})\\
    S_{t+1} &\sim p(.|S_{t},A_{t}), \text{ collect rewards }R_{t} = r(S_{t}, A_{t})\\
    \ldots
  \end{split}
\end{eqnarray*}
Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative reward estimate which takes into account the future. A standard for
modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t.
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following a
policy $\pi$.
\begin{mdframed}
  \textbf{Where is ``value'' buffered in the brain ?}
  The amygdala is known to be involved in significance evaluation) and
  the nucleus accumbens is involved in reward evaluation.
  \end{mdframed}
The  goal of RL is then adapt this policy so as to  maximize the cumulative rewards.
In the definition of $G^\pi$ above, the constant $\gamma$ $(0 \le \gamma < 1$) is the reward \textit{discount factor}.
Taking $\gamma = 0$   corresponds to a perfectly myopic agent who is solely concerned about their
immediate rewards and has no horizon for the future, which is not compatible with long-term goal planning
as implemented in the DMN. For there to be any learning at all, it is thus necessary that $0 < \gamma < 1$.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. The reward importance hence decays exponentially
with distance in the future,
given that there is more uncertainty in the farsighted future.
The reward discount factor $\gamma$ can thus be seen as calibrating
the delay discounting behavior of the intelligent agent.
It is also important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
For instance, a deterministic strategy in playing rock-paper-scissors can
be easily exploited and a uniformly random choice is optimal.

The agent's goal can then be formulated as finding policy $\pi$ which maximizes the
expected cumulative value of state-stat-pairs $(A_0,S_0)$, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|S_0=s,A_0=a,A_1 \sim \pi,A_2 \sim \pi,\ldots].
\end{equation}
That is, the expectated future reward over all possible trajectories in which we start at $s$ take
action $a$ and then follow the policy $\pi$ thereafter to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(S_t,A_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
By construction, maximizing the expectation in  \eqref{eq:qval} is intractable in general
due to the enormous number of possible trajectories the agent might need to
consider.
(If the Bellman-based computational ideas have a corresponding
neurobiological implementation, this provides a framework of how
optimal behavior could be computable in the human brain.)
One popular simplification that removes this issue is to only consider the sub-family of
\textit{deterministic policies}, which map each state to the best single best action which
should be taken at thin state, and therefore define functions from states to actions.
\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} [Sutton and Barto 1998] then takes the simple form
\begin{equation}
  Q^\pi(s, a) = \mathbb E_{s' \sim p(.|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q^\pi(s', a')]
  = \mathbb E_{s' \sim p(.|s,a)} [r(s,a) + \gamma Q^\pi(s', \pi(s'))]
  \label{eq:bellman}
\end{equation}
which provides a recursive decomposition of optimal behavior.
A complicated dynamic programming optimization can be
broken into simpler subproblems at different time points,
which has often been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value,
to guide action (i.e., improve the policy) towards a better state.
It is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

\subsection{Efficient RL with value approximation and replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
A successful RL theory needs ways to approximate the state-action value function
which scale up to large state spaces and / or action spaces.
In state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new state $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$D \leftarrow D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ (hyppocampus ?)
and trying to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q_k)$, with the output of a parametrized regression model $(s,a,r,s')
\mapsto {Q}(s, a|\theta^Q_{k})$, where $Q(.,.|\theta^Q_{k})$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q_{k+1}$.
Computing an optimal policy then corresponds to finding the parameters $\theta^Q_{k+1}$ which minimize the following mean-squared loss function
\begin{equation}
  \mathcal L(\theta^Q_{k+1})
  % = \mathbb E_{(s, a, r, s') \sim \mathcal U(D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q_{k+1}))^2]
  = \mathbb E_{(s, a, r, s') \sim \mathcal U(D)}\left[\frac{1}{2}(Q(s, a|\theta^Q_{k+1}) - y)^2\right],
  \label{eq:oracle}
\end{equation}
where
$ y = r + \gamma \max_{a' \in \mathcal A} Q(s', a'|\theta^Q_k) = r + \gamma Q(s', \pi(s')|\theta^Q_k)$ is the prediction target.
% While $y_t$ is also dependent on $\theta^Q$, this is typically ignored.
For example, a general linear model with a kernel $\phi$ would be of the
form
$${Q}(s, a|\theta^Q) = \phi(s,a)^T\theta^Q.$$
$\phi(s,a)$ would represent a high-level hand-crafted representation of the state-action $(s,a)$ into an appropriate
feature space. Such are the techniques that have been proposed by [sutton et al. xyzw, Song et al. 2016 (NIPS),
  and Parr et al. 2003 (JMLR)].
A breakthrough alternative proposed by Google DeepMind [Minh et al. 2014, Silver et al 2015] is to learn this
representation using a deep neural net, leading to the so-called \textit{Deep Q-learning} family of methods which
are by far the current state-of-the-art in RL. In such a case, $\theta^Q$
would correspond to the strengths of low-level synaptic or high-level functional connections between different neurons / nodes
in the network.

Learning of the the entire model parameters can effictively be achieved via \textit{backprop},
wherein prediction errors are propagated from lower to higher regions to modulate the choice of future
% online stochastic gradient descent is used to minimize the variational objective
% \eqref{eq:oracle}.
